Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-07-21 manuscript/manuscript.Rmd
Continuing use of long parameter names instead of shorthand

raw_data_params <- c("med_cor", "med_SNR")
xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
                 "samps_found", "stans_found", "pooled_cv", "pooled_cv_rob")
all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
                "smp_to_std", "pooled_cv", "pooled_cv_rob", "iso_shape_cor", 
                "iso_area_cor")
```

@@

glmnet_estimates <- both_datasets %>% 
  select(all_of(all_params)) %>% 
  data.matrix() %>%
  glmnet::cv.glmnet(y=both_datasets$feat_bin, family="binomial", alpha=0.5) %>%
  coef() %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  dplyr::rename(`Elastic.net.estimate`=s1)

rf_estimates <- both_datasets %>% 
  select(all_of(all_params)) %>%
  randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
  pluck("importance") %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  select(term, RF.accuracy.decrease=MeanDecreaseAccuracy,
         RF.Gini.Decrease=MeanDecreaseGini)

glm_estimates %>%
  left_join(glmnet_estimates, by="term") %>%
  left_join(rf_estimates, by="term") %>%
  write_csv("manuscript/supplement/supp_table_1_full_model_estimates.csv")
```

We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, the novel peak shape correlation metric `med_cor`, and the novel SNR estimate `med_SNR`, all with reported p-values < 10$^{-10}$. The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp. table 1).

@@

```{r Supp figure param distributions, warning=FALSE}
both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
  bind_rows(.id = "dataset") %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))
pdf(file = "manuscript/supplement/metric_distributions.pdf", width = 6.5, height = 4)
both_datasets %>%
  select(mean_mz:iso_area_cor) %>%
  colnames() %>%
  walk(function(chosen_param){
    hist_gp <- both_datasets %>%
      ggplot() +
      geom_hline(yintercept = 0) +
      geom_histogram(aes(x=get(chosen_param), fill=feat_class), bins=40) +
      scale_fill_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
      labs(y="# of mass features", x=chosen_param) +
      theme_bw() +
      theme(legend.direction = "horizontal", legend.position = "none") +
      facet_wrap(~dataset, ncol=1, scales = "free_y")
    log_gp <- both_datasets %>%
      ggplot(aes(x=get(chosen_param), y=feat_bin)) +
      geom_jitter(aes(color=feat_class), alpha=0.2, height = 0.1, width=0) +
      geom_smooth(method = "glm", method.args=list(family=binomial), formula="y~x", color="black") +
      scale_color_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
      scale_y_continuous(breaks = c(0, 1), labels = c("Bad", "Good"), name=NULL) +
      labs(x=chosen_param) +
      theme_bw() +
      theme(legend.position = "none", 
            axis.text.y = ggtext::element_markdown(
              color = c("#a41118", "#f4bb23"), face = "bold")) +
      facet_wrap(~dataset, ncol=1, scales = "free_y")
    egg::ggarrange(hist_gp, log_gp, nrow=1)
  })
graphics.off()
```

@@

```{r model stability dotplot errorbars supp figure}
both_features <- rbind(FT2040_features, MS3000_features)
dataset_versions_both <- c(dataset_versions, "both_features")
model_param_estimates <- lst(all_params, xcms_params, raw_data_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions_both, function(train_set){
      model_params <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial) %>%
        broom::tidy() %>%
        mutate(train_set)
    }) %>% bind_rows() %>% mutate(param_name)
  }) %>% bind_rows()
suppfig_model_param_estimates_gp <- model_param_estimates %>%
  mutate(param_name=str_remove(param_name, "_params")) %>%
  mutate(param_name=factor(param_name, levels=c("all", "xcms", "raw_data"), 
                           labels=paste(c("Full", "XCMS-only", "Two-parameter"), 
                                        "model"))) %>%
  mutate(train_set=factor(train_set, levels=c("MS3000_features", "FT2040_features", "both_features"),
                          labels=c("MESOSCOPE", "Falkor", "Combined"))) %>%
  ggplot(aes(x=param_name, color=train_set)) +
  geom_hline(yintercept = 0) +
  geom_point(aes(y=estimate), position = position_dodge(width=0.2)) +
  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error),
                position = position_dodge(width=0.2), width=0.3) +
  facet_wrap(~term, scales="free_y", ncol = 4) +
  labs(x="Model type", y="Logistic regression estimate", color="Training set") +
  theme_bw() +
  theme(legend.position = "top", 
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(plot = suppfig_model_param_estimates_gp, 
       filename = "suppfig_model_param_estimates.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 9, units = "in", dpi = 300)
```

@@

gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
  lapply(function(which_params){
    ggplot() +
      geom_hline(yintercept = 0) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
                    ymax=estimate+std.error*2),
                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
                    ymax=estimate+std.error),
                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
      geom_point(aes(x=sample_fraction, y=estimate), 
                 data=sample_model_estimates %>% filter(param_name==which_params),
                 alpha=0.5) +
      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
      theme_bw()
  })
sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
ggsave(plot = sampled_model_est_gp, 
       filename = "sampled_model_est.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

ggsave(plot = fdr_gff_cross_train_gp, 
       filename = "regregrf_fdrgff_cross_train.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 3, units = "in", dpi = 300)
```

>>>>

raw_data_params <- c("med_cor", "med_SNR")
xcms_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", 
                 "sd_pw", "log_mean_area", "log_sd_area", "sn", 
                 "f", "scale", "lmin", "feat_npeaks", "n_found", 
                 "samps_found", "stans_found", "pooled_cv", "pooled_cv_rob")
all_params <- c("mean_mz", "sd_ppm", "mean_rt", "sd_rt", "mean_pw", "sd_pw",
                "log_mean_area", "log_sd_area", "sn", "f", "scale", 
                "lmin", "feat_npeaks", "n_found", "samps_found", "stans_found", 
                "med_cor", "med_SNR", "med_missed_scans", "smp_to_blk", 
                "smp_to_std", "pooled_cv", "pooled_cv_rob", "iso_shape_cor", 
                "iso_area_cor")
param_longnames <- c(
  mean_mz = "Mean feature m/z ratio",
  sd_ppm = "SD of feature m/z ratios in ppm",
  mean_rt = "Mean feature retention time",
  sd_rt = "SD of feature retention time",
  mean_pw = "Mean peak width, in seconds",
  sd_pw = "SD of peak widths, in seconds",
  log_mean_area = "Log-10 scaled mean feature area",
  log_sd_area = "Log-10 scaled SD of feature areas",
  sn = "XCMS 'sn' metric",
  f = "XCMS 'f' metric",
  scale = "XCMS 'scale' metric",
  lmin = "XCMS 'lmin' metric",
  feat_npeaks = "Average number of peaks detected per file",
  n_found = "Proportion of total files in which a peak was found",
  samps_found = "Proportion of total samples in which a peak was found",
  stans_found = "Proportion of total standards in which a peak was found",
  pooled_cv = "Coefficient of variance among pooled sample replicate peak areas",
  pooled_cv_rob = "Robust CV among pooled sample peak areas\n(median absolute deviation divided by median)",
  med_SNR = "Median novel SNR metric across a feature",
  med_cor = "Median peak shape metric across a feature",
  max_SNR = "Maximum novel SNR metric",
  max_cor = "Maximum peak shape metric",
  medtop3_cor = "Median of 3 highest peak shape values",
  log_med_cor = "Median log-10 scaled (1-peak shape)",
  med_missed_scans = "Median number of scans with no data",
  smp_to_blk = "Ratio of average sample peak area to peak area in blank",
  smp_to_std = "Ratio of average sample peak area to peak area in standards",
  iso_shape_cor = "Median log-10 scaled 1-correlation of 13C isotope EIC trace to monoisotopic EIC trace",
  iso_area_cor = "Correlation between 13C isotope area and monoisotopic peak area"
)
```

@@

glmnet_estimates <- both_datasets %>% 
  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
  data.matrix() %>%
  glmnet::cv.glmnet(y=both_datasets$feat_bin, family="binomial", alpha=0.5) %>%
  coef() %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  dplyr::rename(`Elastic.net.estimate`=s1)

rf_estimates <- both_datasets %>% 
  select(all_of(all_params), c("max_SNR", "max_cor", "medtop3_cor", "log_med_cor")) %>% 
  randomForest::randomForest(factor(both_datasets$feat_bin), importance=TRUE) %>%
  pluck("importance") %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  select(term, RF.accuracy.decrease=MeanDecreaseAccuracy,
         RF.Gini.Decrease=MeanDecreaseGini)

param_longnames %>%
  as.data.frame() %>%
  set_names("long_name") %>%
  rownames_to_column("term") %>%
  left_join(glm_estimates, by="term") %>%
  left_join(glmnet_estimates, by="term") %>%
  left_join(rf_estimates, by="term") %>%
  write_csv("manuscript/supplement/supp_table_1_full_model_estimates.csv")
```

We explored the relative predictive power of the individual parameters using the full model and found that the predictors least likely to be different from zero due to chance were the mean *m/z* ratio, our novel peak shape correlation metric, and our novel SNR estimate, all with reported p-values < 10$^{-10}$. The novel parameters were then validated using a random forest model that also found them to have the highest importance (Supp. table 1).

@@

```{r Supp figure param distributions, warning=FALSE}
both_datasets <- list(MESOSCOPE=MS3000_features, Falkor=FT2040_features) %>%
  bind_rows(.id = "dataset") %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_bin=ifelse(feat_class=="Good", 1, 0))

pdf(file = "manuscript/supplement/supp_fig_1_metric_distributions.pdf", width = 6.5, height = 4)
both_datasets %>%
  select(mean_mz:iso_area_cor) %>%
  colnames() %>%
  walk(function(chosen_param){
    hist_gp <- both_datasets %>%
      ggplot() +
      geom_hline(yintercept = 0) +
      geom_histogram(aes(x=get(chosen_param), fill=feat_class), bins=40) +
      scale_fill_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
      labs(y="# of mass features", x=param_longnames[chosen_param]) +
      theme_bw() +
      theme(legend.direction = "horizontal", legend.position = "none",
            axis.title.x = element_text(hjust=0)) +
      facet_wrap(~dataset, ncol=1, scales = "free_y") +
      coord_cartesian(clip = "off")
    log_gp <- both_datasets %>%
      ggplot(aes(x=get(chosen_param), y=feat_bin)) +
      geom_jitter(aes(color=feat_class), alpha=0.2, height = 0.1, width=0) +
      geom_smooth(method = "glm", method.args=list(family=binomial), formula="y~x", color="black") +
      scale_color_manual(breaks = c("Bad", "Good"), values = c("#a41118", "#f4bb23")) +
      scale_y_continuous(breaks = c(0, 1), labels = c("Bad", "Good"), name=NULL) +
      labs(x=NULL) +
      theme_bw() +
      theme(legend.position = "none", 
            axis.text.y = ggtext::element_markdown(
              color = c("#a41118", "#f4bb23"), face = "bold")) +
      facet_wrap(~dataset, ncol=1, scales = "free_y")
    egg::ggarrange(hist_gp, log_gp, nrow=1)
  })
graphics.off()
```

@@

```{r model stability dotplot errorbars supp figure}
both_features <- rbind(FT2040_features, MS3000_features)
dataset_versions_both <- c(dataset_versions, "both_features")
model_param_estimates <- lst(all_params, xcms_params, raw_data_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions_both, function(train_set){
      model_params <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial) %>%
        broom::tidy() %>%
        mutate(train_set)
    }) %>% bind_rows() %>% mutate(param_name)
  }) %>% bind_rows()
suppfig_model_param_estimates_gp <- model_param_estimates %>%
  mutate(param_name=str_remove(param_name, "_params")) %>%
  mutate(param_name=factor(param_name, levels=c("all", "xcms", "raw_data"), 
                           labels=paste(c("Full", "XCMS-only", "Two-parameter"), 
                                        "model"))) %>%
  mutate(train_set=factor(train_set, levels=c("MS3000_features", "FT2040_features", "both_features"),
                          labels=c("MESOSCOPE", "Falkor", "Combined"))) %>%
  ggplot(aes(x=param_name, color=train_set)) +
  geom_hline(yintercept = 0) +
  geom_point(aes(y=estimate), position = position_dodge(width=0.2)) +
  geom_errorbar(aes(ymin=estimate-2*std.error, ymax=estimate+2*std.error),
                position = position_dodge(width=0.2), width=0.3) +
  facet_wrap(~term, scales="free_y", ncol = 4) +
  labs(x="Model type", y="Logistic regression estimate", color="Training set") +
  theme_bw() +
  theme(legend.position = "top", 
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(plot = suppfig_model_param_estimates_gp, 
       filename = "supp_fig_2_model_param_estimates.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 9, units = "in", dpi = 300)
```

@@

gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
  lapply(function(which_params){
    ggplot() +
      geom_hline(yintercept = 0) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
                    ymax=estimate+std.error*2),
                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
                    ymax=estimate+std.error),
                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
      geom_point(aes(x=sample_fraction, y=estimate), 
                 data=sample_model_estimates %>% filter(param_name==which_params),
                 alpha=0.5) +
      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
      theme_bw()
  })
sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
ggsave(plot = sampled_model_est_gp, 
       filename = "supp_fig_3_sampled_model_est.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

ggsave(plot = fdr_gff_cross_train_gp, 
       filename = "supp_fig_4_regregrf_fdrgff_cross_train.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 3, units = "in", dpi = 300)
```
